---
title: "PCA Explorer"
author: "Julien Wist"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_caption: yes
css: visualizeR.css
vignette: >
  %\VignetteIndexEntry{PCA Explorer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Prepare your data

To give an example on how to prepare data, we will use a dataset used in another analysis^[[Comparison of Attenuated Total Reflectance Mid-Infrared, Near Infrared, and 1H-Nuclear Magnetic Resonance Spectroscopies for the Determination of Coffee’s Geographical Origin](https://www.hindawi.com/journals/ijac/2017/7210463/)]. The dataset is provided with the package `visualizeR` and can be called with the `data()` function. 

Since the data have been prepared for another analysis, the data have to be reshaped to be used in the `visualizeR` pipeline. The following elements must be present:

* a vector `ID` of **unique** IDs that is used to identify each entry or sample
* a vector `group` that describes to which class each sample belong
* a table `metadata` of information that belong to each sample and that will be displayed 
* a matrix `x` that contains the data. Each row represents a sample, each columns a variable
* a vector `x_axis` that contains the names or values of each variable

```{r, fig.show='hold'}
library(visualizeR)

data("coffeeNMRSpectra")

ID <- coffeeNMRSpectra$param$catalogID
group <- coffeeNMRSpectra$param$country1
metadata <- data.frame( sapply(coffeeNMRSpectra$param, function(x) x) )
x <- matrix(sapply(I(as.matrix(coffeeNMRSpectra$nmr)), function(x) x), 34, 1610)
x_axis <- as.numeric( colnames(coffeeNMRSpectra$nmr) )

color = sapply(group, function(x) getColor2(as.character(x),'hex'))
```

You can verify that the data includes 34 NMR spectra of Arabica coffee from Brazil, Colombia and Peru. The original data of lenght `r 128*1024` have been binned or bucketed down to 1610 variables.

The `visualizeR` pipeline needs JSON files^[[see](http://www.json.org/)] that look like:

  {'a': 1}
  
where the object contains a key `a` that has a value `1`. JSON are object natively supported in JavaScript, but not very handy in R. In R the best way to represent a JSON is using a `list()`. So each object that must be passed to the `visualizeR` has to be added to a list using the `appendData()` function. Let's create the data object with the following block of commands. 

```{r, fig.show='hold'}
d = list()
c <- data.frame(ID = ID,
                group = group,
                color = color, # can be made faster using levels
                "_highlight" = seq_along(group) - 1,
                data = I(matrix( c(rbind(repRow(x_axis, nrow(x)), x)), nrow(x), ncol(x)*2)),
                I(metadata),
                check.names = FALSE
)
d <- appendData(data = d, variableName = "data", variable = c, type = "table")
```

Note that a vector `_highlight` has been added for internal use. It consist of a unique index and is used to provide highlight of object when hoovered by the mouse in the `visualizer`. Here a simple sequence is used, but in principle any vector of unique IDs could be used. 

Now your data are ready to be pushed to the `visualizer`. 

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
